Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-22 paper.tex
Update tree sequences section

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations.
Through the use of efficient data structures, it is
now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} or
VCF~\citep{danecek2011variant} based formats formats.
This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

\subsection*{Tree sequences}
\label{sec-ts}
One of the key reasons for \msprime's significiant performance advantage
over other simulators is its usage of the ``succinct tree sequence''
(usually, tree sequence) data stucture and the supporting \tskit\
library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we briefly summarise
here.
\jkcomment{We need to mention that tree sequences and tskit are
derived from the 2016 paper and have been extracted from msprime
into a separate C/Python library.}

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} Example tree sequence.
An example tree sequence describing genealogies and genotypes
for four samples at ten sites.
Figure based on~\citet{kelleher2018efficient}.
}
\end{figure}

The succinct tree sequence is a data structure for concisely encoding
genetic ancestry. Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges}. A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
birth population and other metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within a
specific individual, and edges define genetic inheritence
relationships between these nodes. An edge consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the example edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
is a generalisation from the original concept  of
``coalescence records''~\citep{kelleher2016efficient}, and
is the key to the efficiency of tree sequences.
\jkcomment{Summarise and add references?}

The term Ancestral Recombination Graph (ARG) is often understood
to mean the complete record of a sample's genetic
ancestry~\citep[e.g.][]{mathieson2020ancestry}, and in this sense a tree
sequence \emph{is} an ARG. More technically, an ARG is a particular
graphical representation of genetic ancestry based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information---although we do not store
the additional ARG information by default
in \msprime\ because the minimal tree sequence encoding is
much more concise (see the \nameref{sec-arg} section for more details).
It should be noted that, despite efforts to
standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

% JK: commenting this out for now as we're covering it in the Mutations
% section.

% Tree sequences encode a complete and concise description of a set of sampled
% chromosome's genetic ancestry, and they can also efficiently represent
% the genetic variation (extant and ancestral).
% As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
% sequence encoding includes \emph{site} and \emph{mutation} tables. Each
% row in the site table represents some position along the genome, and
% each row in the mutation table represents a change of allelic state
% at a particular site and node. Since the number of mutations is typically
% small, we can usually store the genetic variation for a given variant
% site in constant space. This approach of storing the genetic variation
% via the genetic ancestry is both very concise and extremely efficient
% to process---for example, computing Tajima's $D$ using \tskit\
% is several orders of magnitude faster for large sample sizes than
% efficient libraries working with the variant matrix
% representation~\citep{ralph2020efficiently}.
% \jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
% here. Same basic idea, but much cruder.}

\jkcomment{We need a link paragraph here between what this paper is
about (what we can simulate) and what the rest of this section is about
(efficiently using the \emph{results} of simulations). This is where
we should review the classical approaches. Then lead into the
the discussion of why we need simulations to be efficiently processable,
most importantly, in inference methods like ABC and ML.}

The abilily to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lacking a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output \jkcomment{Also there's the inherent limitation of
newick using branch lenghts; I think there's some citations for
the issues this causes. Probably in the original msprime paper.}.
This has been worked around by using approximate definitions of
concepts like identity by descent (cite carmi paper) \jkcomment{
any other computation methods developed just to work around the
crappiness of ms output?}
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

\jkcomment{A point to make somewhere in here: the tskit data model is
integrated, with all of the information that you need in one file. Not
splattered across a number of different files, that you have to keep track
of.}

\jkcomment{Fit this in somewhere? Or maybe it should be a discussion point?}
Tskit has facilited unprecedented interoperability between simulators:
Forward simulations can also use tree
sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
tree sequences. Msprime can take tree sequences simulated by these forward
simulators as input and complete the simulation (``recapitation'').
See \cite{haller2018tree} for more details.

@@

However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
Others state that an ARG is equivalent to the sequence of marginal
% Also SARGE paper
trees along the genome~\citep{mirzaei2016rent}

@@

>>>>

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations.
Through the use of efficient data structures, it is
now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} or
VCF~\citep{danecek2011variant} based formats.
This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

\subsection*{Tree sequences}
\label{sec-ts}
One of the key reasons for \msprime's significiant performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
(usually, tree sequence) data stucture
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We have subsequently
extracted the tree sequence code from \msprime\ into
\tskit~\citep{tskit2021tskit}, a standalone Python and C library.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
the aspects that are important for simulation here.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} Example tree sequence.
An example tree sequence describing genealogies and genotypes
for four samples at ten sites.
}
\end{figure}

The succinct tree sequence is a data structure for concisely encoding
genetic ancestry and variation. Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
population and other metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within a
some individual, and edges define genetic inheritence
relationships between these nodes. An edge consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the example edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
is a generalisation from the original concept  of
``coalescence records''~\citep{kelleher2016efficient}, and
is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
The term Ancestral Recombination Graph (ARG) is often understood
to mean the complete record of the genetic
ancestry of a set of sampled
genomes~\citep[e.g.][]{mathieson2020ancestry},
and in this sense a tree
sequence \emph{is} an ARG. More technically, an ARG is a particular
graphical representation of genetic ancestry based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although we do not store
the additional ARG information by default
in \msprime\ because the minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section for more details).
It should be noted that, despite efforts to
standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

The point of running simulations is to generate data, and there
are two important ways in which the tree sequences output by
\msprime\ are more useful than classical approaches: (1) tree
sequences are more efficient; and (2) tree sequences contain
more information. We will consider these points in turn.

The classical approach to generating simulation data is
to render the results in some text format. For example,
\ms~\citep{hudson2002generating} outputs a set of sequences
and can also optionally output the marginal trees along
the genome in Newick format,
and variants of this approach are used by many simulators.
Text files have many advantages, but are slow to process
at scale. With sample sizes becoming ever larger, processing
text output can become a significant bottleneck in many analyses.
The ability to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lacking a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
% JK: not sure this is worth leaving in tbh
This lack of reliable information about ancestors in simulation data
has lead, for example, to the definition of segments of
identity by descent as a set of samples having the same time
to their most recent common ancestor along some section of the
genome~\citep{yang2016rapidly}.
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

\jkcomment{Summary paragraph: tskit gives us complete information
about the genetic ancestry. and is very efficient}

% \jkcomment{A point to make somewhere in here: the tskit data model is
% integrated, with all of the information that you need in one file. Not
% splattered across a number of different files, that you have to keep track
% of.}

@@

However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
Others state that an ARG is equivalent to the sequence of marginal
% Also SARGE paper
trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}

@@

Tskit has facilited unprecedented interoperability between simulators:
Forward simulations can also use tree
sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
tree sequences. Msprime can take tree sequences simulated by these forward
simulators as input and complete the simulation (``recapitation'').
See \cite{haller2018tree} for more details.
